Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-04 paper.tex
Finish Edits Section 2,3,4

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} extends intra-workgroup scans to large inputs by resolving inter-workgroup dependencies in three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:}  Each workgroup performs an inclusive scan on its assigned work tile, storing local results and reductions in global memory.
  \item \textbf{Spine Scan:} The workgroup reductions, collectively forming the \emph{spine}, are gathered and scanned to compute the \emph{root}, resolving dependencies between work tiles. For large scans, this stage is an insignificant amount of work compared to the first and third.
  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and adds it to the local scan output.
\end{enumerate}

The absence of inter-workgroup barriers necessitates separate kernel launches for dependency resolution, inherently structuring \emph{Scan-then-Propagate} into three phases. Sengupta et al.'s implementation further complicates this by using recursion to handle arbitrarily large inputs. As the size of a work tile limits the maximum number of elements processed without inter-workgroup dependency resolution, the three-phase structure must be recursively applied until the spine size fits within a single work tile. Thus, an input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$, and $2\cdot\lceil \log_t n \rceil - 1$ kernel launches. Each recursive step $k$ processes $n/t^k$ tiles, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total global data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

@@

Merrill and Grimshaw~\cite{Merrill2009} introduced \emph{workgroup raking} to further reduce recursion overhead when the recursion depth exceeds two. Instead of recursively reducing the spine, the input is partitioned into $o$ groups, where $o$ is the total workgroup occupancy. Each workgroup sequentially processes its assigned partition, reducing the spine size to a constant and guaranteeing a recursion depth of at most two. This reduces global memory movement for the spine scan from $O(n/t)$ to $O(2o)$ and limits kernel launches to three, reducing total global memory movement to $O(3n + 4o) = O(3n)$.

Breitbart~\cite{10.5555/2031978.2032029} applied Gupta et al.'s \emph{Persistent-Threads}~\cite{gupta2012} to scan, eliminating kernel launch overheads. Recognizing that raking naturally aligns with this paradigm, Breitbart replaced kernel launches with an atomics-based inter-workgroup barrier. Since multiprocessors are not oversubscribed in raking, workgroups persist for the entire kernel duration, allowing synchronization at the barrier without context switching. Despite maintaining $O(3n)$ global memory movement, consolidating the scan into a single kernel minimizes launch overheads.

@@

\subsection{Ideal Scan Architecture}
Beyond their differences in speed, \emph{Chained-Scan} and \emph{Reduce-Then-Scan} embody two fundamentally contrasting paradigms. By eliding explicit inter-workgroup synchronization, \emph{Chained-Scan} aligns naturally with the oversubscription paradigm, delegating occupancy management to the hardware scheduler. In contrast, the raking approach inherent to \emph{Reduce-Then-Scan} mandates the fixed-launch paradigm, necessitating explicit control over occupancy and potentially synchronization.

Fixed-launch approaches demand substantial developer effort, requiring either tuning of workgroup launch sizes across target devices or occupancy discovery. Moreover, previous workgroup occupancy discovery techniques, such as the one proposed by Sorensen et al.~\cite{sorensen2016}, are infeasible to implement in WebGPU due to their reliance on memory fences---a feature not supported by the specification.\footnote{Our implementation includes a fenceless occupancy discovery method amenable to the WebGPU environment, but its description is beyond the scope of this work.}

@@

Thus, oversubscription and \emph{Chained-Scan} emerges as the preferred architecture, as it naturally aligns with portability by leveraging hardware schedulers to dynamically manage occupancy across diverse architectures without requiring developer intervention. Furthermore, as a single kernel with inherently serial ordering, \emph{Chained-Scan} is suitable for \emph{in-situ} operations, augmenting practical routines such as data deduplication, sparse data compaction, error correction, and interval merging.

But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in terms of its theoretical and implementation characteristics, it falls short in fully meeting the criterion of \emph{portability}. This limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA devices but notably lacking on Apple's M-series and ARM GPUs~\cite{sorensen2021}. Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure in the form of deadlocking.

@@

The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}. Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to \emph{false-positive} fallback operations. As a result, memory bandwidth usage increases (and throughput decreases) from $O(fn)$ to $O((2 + f)n)$. However, as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache. As we later empirically demonstrate, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile. Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.

>>>>

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} extends intra-workgroup scans to large inputs by resolving inter-workgroup dependencies in three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:}  Each workgroup performs an inclusive scan on its assigned work tile, storing local results and reductions in global memory.
  \item \textbf{Spine Scan:} The workgroup reductions, collectively forming the \emph{spine}, are gathered and scanned to compute the \emph{root}, resolving dependencies between work tiles. For large scans, this stage is an insignificant amount of work compared to the first and third.
  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and adds it to the local scan output.
\end{enumerate}
The absence of inter-workgroup barriers necessitates separate kernel launches for dependency resolution, naturally enforcing this structure. To accommodate arbitrarily large inputs, Sengupta et al. employ recursion, repeatedly applying the three-phase process until the spine fits within a single work tile. Thus, an input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$, and $2\cdot\lceil \log_t n \rceil - 1$ kernel launches. Each recursive step $k$ processes $n/t^k$ tiles, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total global data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

@@

Merrill and Grimshaw~\cite{Merrill2009} introduced \emph{workgroup raking} to reduce recursion overhead for arbitrarily large inputs. Instead of recursively reducing the spine, the input is partitioned among a fixed number of workgroups, $c$, where $c \geq o$, the total workgroup occupancy. This constrains the spine from a variable, potentially recursive structure to a fixed size $c$, ensuring a recursion depth of at most two. As a result, global memory movement for the spine scan is reduced from $O(n/t)$ to $O(2c)$, and kernel launches are limited to three, yielding a total global memory movement of $O(3n + 4c) = O(3n)$.

Breitbart~\cite{10.5555/2031978.2032029} applied Gupta et al.'s \emph{Persistent-Threads}~\cite{gupta2012} to scan, eliminating kernel launch overheads. Recognizing that raking naturally aligns with this paradigm, Breitbart's method dynamically \emph{discovers} the workgroup occupancy and launches exactly $o$ kernels. Since multiprocessors are not oversubscribed, workgroups persist for the entire kernel duration, enabling synchronization at an atomics-based inter-workgroup barrier without context switching. While global memory movement remains $O(3n)$, this consolidates the scan into a single kernel, minimizing launch overheads.

@@

\subsection{Ideal Scan Architecture}
\emph{Chained-Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design: one embraces oversubscription and dynamic scheduling, while the other relies on a fixed-launch model with explicit control over occupancy. By avoiding explicit inter-workgroup synchronization, \emph{Chained-Scan} naturally aligns with the oversubscription paradigm, allowing the hardware scheduler to dynamically manage occupancy. In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, which mandates explicit occupancy control and, in some cases, inter-workgroup synchronization.

Fixed-launch approaches, such as those used in \emph{Reduce-Then-Scan}, introduce several challenges. They require either careful tuning of workgroup launch sizes across different architectures or an occupancy discovery mechanism, which is nontrivial to implement. For example, the workgroup occupancy discovery technique proposed by Sorensen et al.~\cite{sorensen2016} relies on memory fences—unsupported in WebGPU—making it infeasible in certain environments.\footnote{Our implementation includes a fenceless occupancy discovery method suitable for WebGPU, but its description is beyond the scope of this work.}

@@

Given these constraints, oversubscription and \emph{Chained-Scan} emerge as the more portable and adaptable architecture. By leveraging the hardware scheduler, \emph{Chained-Scan} dynamically distributes work across architectures without requiring developer intervention. Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging.

However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series and certain ARM GPUs do not. Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient—it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

@@

The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}. Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to \emph{false-positive} fallback operations. As a result, memory bandwidth usage increases (and we expect throughput to decrease) by $O(fn)$ to $O((2 + f)n)$. However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache. As we later empirically demonstrate(Simulated Blocking Section), this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile. Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.
